---
title: "Storm Data Analysis - Peer Assessment 2"
author: ""
date: "July 25, 2014"
output: html_document
---


## Synopsis

This data analysis uses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to address the following questions:

1.  Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

The purpose of this analysis is to review historical severe weather events in order to prioritize resources for future contingencies.



## Data Processing

### Download storm data, documentation and FAQ

Set working directory:
``` {r setwd, echo = TRUE}
setwd("~/GitHub/RepData_PeerAssessment2")
```
Create a data subdirectory if it does not exist:
``` {r subdirectory, echo = TRUE} 
if (!file.exists("data")) {
        dir.create("data")
}
```
Download *Storm Data* and unzip if it isn't already there:
``` {r download dataset, echo = TRUE}
library(R.utils, quietly = TRUE)                        #for bzunzip
fileDir <- "./data/"
fileZipName <- "repdata-data-StormData.csv.bz2"
fileName <- "repdata-data-StormData.csv"
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

filePath <- paste0(fileDir, fileZipName)
if (!file.exists(filePath)) {
        download.file(fileUrl, destfile = filePath, method ="curl")
        bunzip2(filePath, exdir=fileDir, remove=FALSE)
        dateDownloaded <- date()
}
```
Download documentation files:  *National Weather Service Storm Data Documentation* and *National Climatic Data Center Storm Events FAQ*
```{r download documentation, echo = TRUE}
fileName2 <- "repdata-peer2_doc-pd01016005curr.pdf"
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
filePath2 <- paste0(fileDir, fileName2) 
if (!file.exists(filePath2)) {
        download.file(fileUrl2, destfile = filePath2, method ="curl")
        dateDownloaded <- date()
}

fileName3 <- "repdata-peer2_doc-NCDC Storm Events-FAQ Page.pdf"
fileUrl3 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
filePath3 <- paste0(fileDir, fileName3) 
if (!file.exists(filePath3)) {
        download.file(fileUrl3, destfile = filePath3, method ="curl")
        dateDownloaded <- date()
}
```

### Read in data file, subset and clean
``` {r read dataset, echo = TRUE}
raw <- read.csv(paste0(fileDir, fileName))
data <- subset(raw, na.omit=TRUE, select=c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP))
```

### Run analysis

Compute property damage cost by multiplying damage variable *PROPDMG* by the exponent variable *PROPDMGEXP*
``` {r exponents, echo = TRUE}
cost <- rep(NA, nrow(data))                                                     # damage cost
for (i in 1:nrow(data)) {                               
        if (sum(data[i,"PROPDMGEXP"] == c(0:9)) != 0) {                         # exponent 0-9
                cost[i] <- data[i,"PROPDMG"] * 
                        10^as.integer(as.character(data[i,"PROPDMGEXP"]))
        }
        else if (sum(data[i,"PROPDMGEXP"] == c("b","B")) != 0) {                # billions        
                cost[i] <- data[i,"PROPDMG"] * 10^9
        }
        else if (sum(data[i,"PROPDMGEXP"] == c("m","M")) != 0) {                # millions
                cost[i] <- data[i,"PROPDMG"] * 10^6
        }
        else if (sum(data[i,"PROPDMGEXP"] == c("h","H","k","K")) != 0) {        # thousands
                cost[i] <- data[i,"PROPDMG"] * 10^3
        }        
        else {
                cost[i] <- data[i,"PROPDMG"]
        }
}
data <- cbind(data,cost)
```
Determine total injuries, deaths and property damage costs by weather event type:
```{r analyze, echo = TRUE}
toplevels <- 20
unsorted_injured_by_event <- tapply(data$INJURIES, data$EVTYPE, 
                           sum, simplify = TRUE)                # simply returns a scalar not list
injured_by_event <- sort(unsorted_injured_by_event, decreasing = TRUE)
top_injury_events <- head(injured_by_event, n=toplevels)
top_injury_events <- sort(top_injury_events, decreasing = FALSE)        # reverse sort order

unsorted_deaths_by_event <- tapply(data$FATALITIES, data$EVTYPE, sum, simplify = TRUE)
deaths_by_event <- sort(unsorted_deaths_by_event, decreasing = TRUE)
top_death_events <- head(deaths_by_event, n=toplevels)
top_death_events <- sort(top_death_events, decreasing = FALSE)                  

unsorted_cost_by_event <- tapply(data$cost, data$EVTYPE, sum, simplify = TRUE)
cost_by_event <- sort(unsorted_cost_by_event, decreasing = TRUE)
top_cost_events <- head(cost_by_event, n=toplevels)
top_cost_events <- sort(top_cost_events, decreasing = FALSE)                  
```
Tabulate the results for deaths, injuries and damage costs:
```{r }
combo <- cbind(unsorted_deaths_by_event, unsorted_injured_by_event, 
               as.integer(unsorted_cost_by_event/10^6))
results <- combo[order(-unsorted_deaths_by_event),]
colnames(results) <- c("Deaths","Injuries","Damage($M)")
print(head(results, n=toplevels))
```


## Results

With respect to analyzing the question *which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?*, the plots in Figures 1 & 2 are illustrative.
```{r figure1, echo = TRUE}
library(lattice)
barchart(top_death_events, col="red", xlab='Deaths', main='Figure 1 Total Deaths by Weather Event')
```

```{r figure2, echo = TRUE}
barchart(top_injury_events, xlab='Injuries', main='Figure 2 Total Injuries by Weather Event')
```


In terms of studying the question *which types of events have the greatest economic consequences?*
the following plot is informative.
```{r figure3, echo=TRUE}
barchart(top_cost_events, col="green", xlab='Cost ($US)', 
         main='Figure 3 Total Property Damage by Weather Event')
```

