---
title: "Storm Data Analysis - Peer Assessment 2"
author: ""
date: "July 20, 2014"
output: html_document
---
## Synopsis
 >>>describes and summarizes your analysis in at most 10 complete sentences.

This data analysis uses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. 
to address the following questions:

1.  Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing

### Download storm data, documentation and FAQ

Set working directory:
``` {r setwd }
setwd("~/GitHub/RepData_PeerAssessment2")
```
Create a data subdirectory if it does not exist:
``` {r subdirectory} 
if (!file.exists("data")) {
        dir.create("data")
}
```
Download *Storm Data* and unzip if it isn't already there:
``` {r download dataset}
library(R.utils)       #for bzunzip
fileDir <- "./data/"
fileZipName <- "repdata-data-StormData.csv.bz2"
fileName <- "repdata-data-StormData.csv"
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

filePath <- paste0(fileDir, fileZipName)
if (!file.exists(filePath)) {
        download.file(fileUrl, destfile = filePath, method ="curl")
        bunzip2(filePath, exdir=fileDir)
        dateDownloaded <- date()
}
```
Download documentation files:  *National Weather Service Storm Data Documentation* and *National Climatic Data Center Storm Events FAQ*
```{r download documentation}
fileName2 <- "repdata-peer2_doc-pd01016005curr.pdf"
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
filePath2 <- paste0(fileDir, fileName2) 
if (!file.exists(filePath2)) {
        download.file(fileUrl2, destfile = filePath2, method ="curl")
        dateDownloaded <- date()
}

fileName3 <- "repdata-peer2_doc-NCDC Storm Events-FAQ Page.pdf"
fileUrl3 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
filePath3 <- paste0(fileDir, fileName3) 
if (!file.exists(filePath3)) {
        download.file(fileUrl3, destfile = filePath3, method ="curl")
        dateDownloaded <- date()
}
```

### Read in data file
2. Process/transform the data into a format suitable for analysis
``` {r }
raw <- read.csv(paste0(fileDir, fileName))
# raw <- read.csv(paste0(fileDir, fileName),colClasses = c("integer", "character", "integer"))
#clean <- na.omit(raw)
#clean$date <- as.factor(clean$date)
#clean$interval <- as.factor(clean$interval)
```

## Results
You may have other sections in your analysis, but Data Processing and Results are required.

The analysis document must have at least one figure containing a plot.

Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default setting in knitr).


```{r }

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
